---
title: "Extending Hypothesis Testing to Multiple Comparisons"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"
format: pptx

execute:
  echo: true
  warning: false
  message: false
---

### Module Review

::: {style="font-size:0.6em"}

`hypothesis testing` : statistical approach used to evaluate whether sample data provide sufficient evidence against a claim about a population

General hypothesis-testing framework:

- defining a population parameter of interest
- stating null and alternative hypotheses
- calculating a test statistic
- determining a P-value
- evaluating evidence against the null hypothesis
- drawing a conclusion in context

Hypothesis testing focuses on evaluating one research question at a time

Examples:

- Is a population mean different from a benchmark?
- Are two group means different?
- Are two population proportions different?
- Is a predictor associated with an outcome?

:::

---

### Module Review

::: {style="font-size:0.6em"}

Biomedical inquiries often involve evaluating multiple feature simultaneously - biomarkers, clinical variables, treatment effects, genetic factors, or regression coefficients - within the same study

For example, a cancer study may simultaneously evaluate whether:

- tumor stage predicts survival
- age predicts survival
- mutation status predicts survival
- treatment exposure predicts survival

These studies often generate many hypothesis tests rather than a single test

When numerous tests are performed, statistically significant results may occur by chance alone

Multiple hypothesis testing focuses on interpreting results across many statistical tests

:::

---

### Definition: Multiple Hypothesis Testing

::: {style="font-size:0.6em"}

`multiple hypothesis testing` : process of evaluating two or more statistical hypotheses within a single study or analysis

- when number of hypothesis tests increases $\rightarrow$ probability of obtaining false-positive [Type I error rate] results by chance increases

Examples include:

- testing hundreds of genes for differential expression
- testing thousands of genomic variants
- evaluating many biomarkers simultaneously
- screening multiple clinical outcomes

Each test produces its own:

- null hypothesis
- test statistic
- p-value : evaluation of hundreds or thousands of p-values simultaneously

Note: probability of obtaining false positive findings increases as the number of tests increases

:::
---

### Why Multiple Testing Procedures Are Needed

::: {style="font-size:0.6em"}

- hypothesis testing is designed to evaluate evidence against a null hypothesis

- when a single hypothesis test is performed, the probability of a false positive is controlled by the selected significance level: α

- when many hypothesis tests are performed simultaneously, opportunities for false positive findings increase

As a result:

- statistically significant findings may occur by chance alone
- number of false discoveries increases
- scientific conclusions become less reliable
- follow-up studies may fail to replicate reported findings

Multiple-testing procedures extend the hypothesis-testing framework to settings involving many simultaneous tests.

Multiple-testing involves two key goals:

- identify true associations
- limit false positive findings


:::
---

### Motivation for Multiple Testing Corrections

::: {style="font-size:0.76em"}

Consider: to understand why multiple testing can lead to false positives, consider a situation where all null hypotheses are actually true

If α=0.05 then each individual hypothesis test has a 5% probability of producing a false positive result

When many tests are performed false positives accumulate

Examples:

| Number of Tests | Expected False Positives |
|-----------------|--------------------------|
| 10              | 0.5                      |
| 100             | 5                        |
| 1000            | 50                       |

Note : even when no true associations exist, statistically significant results can still appear

:::
---

### Goals and Approaches for Multiple Hypothesis Testing

::: {style="font-size:0.76em"}

Multiple-testing procedures attempt to balance two competing objectives:

**1. Identify true biological signals**

Discovering meaningful associations that may improve scientific understanding

**2. Limit false positive findings**

Avoid reporting associations that arise only from random variation

:::
---



### Goals and Approaches for Multiple Hypothesis Testing

::: {style="font-size:0.76em"}

Two Approaches for Multiple Testing:

`Family-Wise Error Rate [FWER]`: method that controls the probability of making at least one false positive across all tests

Method: Bonferroni correction

`False Discovery Rate [FDR]` : controls the expected proportion of false positives among significant findings

Method: Benjamini-Hochberg procedure

:::
---

### When to Apply Multiple Hypothesis Corrections

::: {style="font-size:0.6em"}

Consider:

> Are multiple hypothesis tests being interpreted as part of the same study or analysis?

Multiple hypothesis correction should be applied whenever several p-values are generated from related questions

Apply approach if repeated testing increases the chance of false discoveries

Examples:

- 5–20 related tests: multiple-testing correction should usually be considered
- dozens to thousands of tests: correction is generally expected
- high-dimensional omics studies: FDR correction is standard

:::

---

### Controlling False Positives in Multiple Hypothesis Testing

::: {style="font-size:0.6em"}

When many hypothesis tests are performed, one can decide how they want to control false positives

| Question | FWER | FDR |
|:----------|:----------|:----------|
| What is controlled? | Probability of making one or more false positive conclusions anywhere in the entire study | Proportion of significant findings that are expected to be false positives |
| Common method | Bonferroni correction | Benjamini-Hochberg procedure |
| Stringency | Very conservative | Less conservative |
| Statistical power | Lower | Higher |
| Best for | Confirmatory analyses | Exploratory analyses |


:::
---

#### Controlling False Positives in Multiple Hypothesis Testing

::: {style="font-size:0.5em"}

**Family-Wise Error Rate [FWER]**

- when single false positive would be costly
- results may directly influence clinical decisions
- regulatory or confirmatory studies are being conducted
- number of hypotheses is relatively small

Examples

- Phase III clinical trials
- Primary endpoints in biomedical studies
- regulatory submission analyses



:::
---

#### Controlling False Positives in Multiple Hypothesis Testing

::: {style="font-size:0.5em"}


**False Discovery Rate [FDR]**

- many hypotheses are tested simultaneously
- some false positives are acceptable in exchange for increased discovery
- findings will be validated in future studies

Examples

- gene expression studies
- GWAS analyses
- proteomics and other omics studies
- biomarker discovery projects

:::
---


### Example 1: Multiple Clinical Outcomes

::: {style="font-size:0.72em"}

Background:

A randomized cancer therapy study evaluates whether a new treatment is associated with improvement across six clinical outcomes.

Each outcome represents a separate hypothesis test.

For each outcome:

$$
H_0: \text{new treatment is not associated with the outcome}
$$

$$
H_A: \text{new treatment is associated with the outcome}
$$

Scientific question:

> Which clinical outcomes remain signficant after accounting for the fact that several hypothesis tests were performed?

:::

---

### Example 1: What Tests Generated These P-values?

::: {style="font-size:0.6em"}

Each p-value comes from a separate hypothesis test evaluating whether treatment is associated with a patient outcome

Possible tests include:

| Outcome | Example hypothesis test |
|---|---|
| Response Rate | logistic regression |
| Symptom Score | Two-sample t test|
| Quality of Life | linear regression |
| Hospitalizations | Poisson regression|
| Overall Survival | Cox regression or log-rank test |
| Progression-Free Survival | Cox regression or log-rank test |

Each test has its own null hypothesis for example:

$$
H_0: \text{treatment is not associated with the outcome}
$$

- all six p-values are interpreted together as evidence about treatment benefit: multiple-testing correction is needed

:::

---

### Example 1: Simulated Clinical Endpoint Results

::: {style="font-size:0.6em"}
```{r}
library(tidyverse)

set.seed(300)

clinical_df <- tibble(
  outcome = c(
    "Overall Survival",
    "Progression-Free Survival",
    "Response Rate",
    "Symptom Score",
    "Quality of Life",
    "Hospitalizations"
  ),
  p_value = c(0.002, 0.018, 0.031, 0.041, 0.12, 0.21)
)

clinical_df
```


**Output Overview**

- table of raw p-values from six hypothesis tests

- each p-value represents evidence against a separate null hypothesis

- raw p-values treats each test as if it were the only test performed

:::

---

### Apply Bonferroni and FDR Correction

::: {style="font-size:0.55em"}

```{r, echo=TRUE}
clinical_df |>
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    p_fdr = p.adjust(p_value, method = "fdr"),
    raw_significant = p_value <= 0.05,
    bonferroni_significant = p_bonferroni <= 0.05,
    fdr_significant = p_fdr <= 0.05
  )


```


**Code Interpretation**

`Bonferroni correction` : controls the probability of making at least one false positive across all six tests

`FDR correction`: controls the expected proportion of false positives among the outcomes declared significant

`Central Aim` : Determine whether results remain statistically credible after accounting for all related tests?

- `p_bonferroni`: p-values adjusted to control the probability of at least one false positive
- `p_fdr`: p-values adjusted using the Benjamini-Hochberg false discovery rate procedure
- `raw_significant`: significant before correction
- `bonferroni_significant`: significant after Bonferroni correction
- `fdr_significant`: significant after FDR correction

:::

---


### Apply Bonferroni and FDR Correction

::: {style="font-size:0.55em"}

```{r, echo=TRUE}
clinical_df |>
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    p_fdr = p.adjust(p_value, method = "fdr"),
    raw_significant = p_value <= 0.05,
    bonferroni_significant = p_bonferroni <= 0.05,
    fdr_significant = p_fdr <= 0.05
  )


```


**Interpretation**

Raw p-value interpretation - four outcomes appeared statistically significant:

- Overall Survival
- Progression-Free Survival
- Response Rate
- Symptom Score

:::

---

### Apply Bonferroni and FDR Correction

::: {style="font-size:0.55em"}

```{r, echo=TRUE}
clinical_df |>
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    p_fdr = p.adjust(p_value, method = "fdr"),
    raw_significant = p_value <= 0.05,
    bonferroni_significant = p_bonferroni <= 0.05,
    fdr_significant = p_fdr <= 0.05
  )


```


**Interpretation**

After applying Bonferroni and FDR correction, only Overall Survival remained statistically significant at: α=0.05

- some significant findings observed in the raw analysis were no longer statistically significant after accounting for multiple hypothesis tests
- overall Survival showed the strongest evidence against the null hypothesis
- multiple-testing procedures help distinguish findings that remain credible after considering the full set of hypothesis tests performed

:::

---

### Example 2: Cancer Biomarker Discovery Screen

::: {style="font-size:0.6em"}

Background:

A cancer biomarker discovery study evaluates whether baseline tumor biomarkers are associated with treatment response

The study tests 1,000 biomarkers and 1,000 hypothesis tests generated

**For each biomarker**:

$H_{0j}$ : biomarker j is not associated with treatment response

$H_{Aj}$ : biomarker j is associated with treatment response


- each test is evaluated using: α=0.05

- many biomarkers may appear significant by chance alone

:::

---

### Simulate Biomarker P-values

::: {style="font-size:0.6em"}
```{r}
set.seed(160)

n_tests <- 1000
n_true_signals <- 50
n_null <- n_tests - n_true_signals

biomarker_results <- tibble(
  biomarker = paste0("BMK", stringr::str_pad(1:n_tests, 4, pad = "0")),
  truth = c(
    rep("True association", n_true_signals),
    rep("No association", n_null)
  ),
  p_value = c(
    rbeta(n_true_signals, shape1 = 0.35, shape2 = 8),
    runif(n_null, min = 0, max = 1)
  )
)

biomarker_results |>
  slice_head(n = 10)
```

- simulated dataset contains two types of biomarkers

- some biomarkers have true associations [p-values concentrated near zero]

- most biomarkers have no true association [p-value - distribution between 0 and 1]

- small subset of biomarkers may represent real biological signal

:::
---

### Raw P-value Results

```{r}
raw_results <- biomarker_results |>
  mutate(raw_significant = p_value <= 0.05) |>
  count(truth, raw_significant)

raw_results
```


::: {style="font-size:0.68em"}

**Table Summary**

- count of biomarkers significant based on unadjusted p-values

- note number of biomarkers with no association that appear significant [n=43] - false positives

- false positives generated because many hypothesis tests were performed

:::
---

### Bonferroni and FDR Adjustments

::: {style="font-size:0.58em"}

```{r, echo=TRUE}
biomarker_adjusted <- biomarker_results |>
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    p_fdr = p.adjust(p_value, method = "fdr"),
    raw_significant = p_value <= 0.05,
    bonferroni_significant = p_bonferroni <= 0.05,
    fdr_significant = p_fdr <= 0.05
  )

```

**Code Overview**

Compute Bonferroni-adjusted p-values: `p_bonferroni = p.adjust(p_value, method = "bonferroni")`

Computes Benjamini-Hochberg adjusted p-values: `p_fdr = p.adjust(p_value, method = "fdr")`

Identify biomarkers significant using unadjusted p-values: `raw_significant = p_value <= 0.05`

Identify biomarkers significant after Bonferroni correction: `bonferroni_significant = p_bonferroni <= 0.05`

Identify biomarkers significant after FDR correction: `fdr_significant = p_fdr <= 0.05`

:::
---

### Bonferroni and FDR Adjustments

::: {style="font-size:0.58em"}


```{r}
biomarker_adjusted <- biomarker_results |>
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    p_fdr = p.adjust(p_value, method = "fdr"),
    raw_significant = p_value <= 0.05,
    bonferroni_significant = p_bonferroni <= 0.05,
    fdr_significant = p_fdr <= 0.05
  )

biomarker_adjusted |>
  summarise(
    raw_discoveries = sum(raw_significant),
    bonferroni_discoveries = sum(bonferroni_significant),
    fdr_discoveries = sum(fdr_significant)
  )
```

Table contains total count of statistically significant biomarkers

- `raw_discoveries`: biomarkers significant using unadjusted p-values
- `bonferroni_discoveries`: biomarkers significant after strict FWER control
- `fdr_discoveries`: biomarkers significant after FDR control

:::
---

### Bonferroni and FDR Adjustments

::: {style="font-size:0.58em"}


```{r}
biomarker_adjusted <- biomarker_results |>
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    p_fdr = p.adjust(p_value, method = "fdr"),
    raw_significant = p_value <= 0.05,
    bonferroni_significant = p_bonferroni <= 0.05,
    fdr_significant = p_fdr <= 0.05
  )

biomarker_adjusted |>
  summarise(
    raw_discoveries = sum(raw_significant),
    bonferroni_discoveries = sum(bonferroni_significant),
    fdr_discoveries = sum(fdr_significant)
  )
```

**Results Interpretation**

- 79 biomarkers had unadjusted p-values ≤ 0.05 [many might be false positives]

- 4 biomarkers remained significant after Bonferroni correction

  - Bonferroni tries to prevent even a single false positive across all tests
  
- 9 biomarkers remained significant after FDR correction

  - FDR controls the expected proportion of false positives rather than eliminating them entirely

:::
---

### Summary

::: {style="font-size:0.72em"}

- multiple hypothesis testing is implemented when many statistical tests are evaluated within the same study

- each hypothesis test produces its own p-value
- performing many tests increases the chance of false positive findings
- raw p-values can overstate evidence when many hypotheses are evaluated simultaneously
multiple-testing procedures help control false discoveries

Two hypothesis correction procedures include:

Family-Wise Error Rate [FWER]/Bonferroni Correction: controls the probability of making at least one false positive

False Discovery Rate [FDR]/Benjamini-Hochberg Correction:	controls the expected proportion of false positives among significant findings

- multiple hypothesis testing correction helps ensure that statistically significant findings remain scientifically credible when many hypotheses are evaluated simultaneously

:::

